---
title: "Unsupervised Exploratory Analysis of TCGA BRCA Patient Segments"
author: "Tejaswi S Iyyanki"
date: "2/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r silent calling libraries, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
require("preprocessCore")
library(TCGAbiolinks)
library(SummarizedExperiment)

#library(devtools)
library(ggfortify)
library(ggplot2)
library(ConsensusClusterPlus)
```


```{r Open data and process, echo=FALSE, results='hide', message=FALSE, warning=FALSE}

geneExp_T=read.csv(file = "../TCGA_BRCA_Tumor_RSEM_GeneExp.tsv", header = T, sep = "\t", as.is = T, row.names = NULL)

geneExp_T[1:5,1:5]

geneExp_T_m=as.matrix(geneExp_T[,2:ncol(geneExp_T)])
rownames(geneExp_T_m)=geneExp_T$row.names

#Now it is a matrix format
geneExp_T_m[1:5,1:5]

#Size of matrix
dim(geneExp_T_m)

#Filter out Protein-coding genes
ENSEMBL_annotations=read.csv("Homo_sapiens.GRCh37.87_Annotations.txt",
                             header=FALSE,sep=" ",stringsAsFactors = F)
ENSEMBL_annotations_prot=unique(ENSEMBL_annotations[ENSEMBL_annotations$V3=="protein_coding",])

geneExp_T_m_prot=subset(geneExp_T_m,rownames(geneExp_T_m) %in% ENSEMBL_annotations_prot$V2)
#Now we have gene matrix (rows) consisting of protein-coding ONLY
dim(geneExp_T_m_prot)
head(geneExp_T_m_prot)[,1:5]

```


```{r Filter Data, echo=FALSE, results='hide', message=FALSE, warning=FALSE}

#Remove <2 TPM genes; Only retain rows (genes) that show atleast 20% of columns (patients) have TPM values > 2.
TCGA_only_f1=geneExp_T_m_prot
TCGA_only_f1=geneExp_T_m_prot[(rowSums(geneExp_T_m_prot>2)>=as.integer(0.2*ncol(geneExp_T_m_prot))),] 
dim(TCGA_only_f1) 
TCGA_only_f1[1:5,1:5]

#Quantile Normalize data
TCGA_only_f1_prot_QN=normalize.quantiles(TCGA_only_f1)
row.names(TCGA_only_f1_prot_QN)=row.names(TCGA_only_f1)
colnames(TCGA_only_f1_prot_QN)=colnames(TCGA_only_f1)
TCGA_only_f1_prot_QN[1:5,1:5]

#Log2 transform data
TCGA_only_f1_prot_QN_l2=log2(TCGA_only_f1_prot_QN+1)
TCGA_only_f1_prot_QN_l2[1:5,1:5]

#Mean-normalize matrix
TCGA_only_f1_prot_QN_l2_m=TCGA_only_f1_prot_QN_l2-apply(TCGA_only_f1_prot_QN_l2,1,mean)
TCGA_only_f1_prot_QN_l2_m[1:5,1:5]

#Get top5K variable genes
var_TCGA_only=apply(TCGA_only_f1_prot_QN_l2, 1, var)
TCGA_only_f1_prot_QN_l2_5K=TCGA_only_f1_prot_QN_l2[order(var_TCGA_only, decreasing=TRUE)[1:5000],]
TCGA_only_f1_prot_QN_l2_m_5K=TCGA_only_f1_prot_QN_l2_m[order(var_TCGA_only, decreasing=TRUE)[1:5000],]

#Most variable genes (top 25)
sort(var_TCGA_only, decreasing=TRUE)[1:25]
```

*Supervised k-means clustering is used here. Again the goal is NOT to determine the number of row or column clusters here.*

```{r visualization Heatmap, echo=TRUE}

#Use the top5K features (genes) which normalized by its mean value for each row and visualize as a Heatmap
#Kmeans here
Heatmap(TCGA_only_f1_prot_QN_l2_m_5K,
        col = colorRamp2(c(-2, 0, 2),c("purple", "#EEEEEE", "yellow")),
        show_column_names = F, show_row_names = F,
        cluster_columns = T,cluster_rows = T,
        show_column_dend = T, show_row_dend = T,
        clustering_distance_rows = "pearson", clustering_distance_columns = "pearson",
        show_heatmap_legend =T,
        row_km = 8, column_km = 4
        )
```

*Unsupervised Hierarchical clustering is used here*
``` {r Hierarchical, echo=TRUE}

#Hierarchical and unsupervised clustering
Heatmap(TCGA_only_f1_prot_QN_l2_m_5K,
        col = colorRamp2(c(-2, 0, 2),c("blue", "#EEEEEE", "orange")),
        show_column_names = F, show_row_names = F,
        cluster_columns = T,cluster_rows = T,
        show_column_dend = T, show_row_dend = T,
        clustering_distance_rows = "pearson", clustering_distance_columns = "pearson",
        show_heatmap_legend =T
        )

```

## Kmeans clustering and PCA plot to visualize patient segments 

```{r PCA,echo=TRUE}


PLOTPCA_PRCOMP=function(inputM,clus) {
    data_in=t(
                inputM
                )
    res.pca=prcomp(data_in,
                    center = TRUE,scale. = TRUE)
    df_data_in=as.data.frame(data_in)
    df_data_in$clus=clus
    #Now plot PCA
    autoplot(res.pca,data =df_data_in,colour="clus",label=F,title=paste(topNgenes))
}

#As you can see colnames have changed from "-" to "."
colnames(TCGA_only_f1_prot_QN_l2_5K)[1:5]

#So convert this according to this code:
colnms=unlist(lapply(colnames(TCGA_only_f1_prot_QN_l2_5K),function(x) paste((strsplit(x, "[.]"))[[1]],collapse="-") ) )
colnms[1:5]

dtm=TCGA_only_f1_prot_QN_l2_5K
colnames(dtm)=colnms
        
KM_clus_cols=kmeans(t(dtm), 4)

#Notice that we are using 5,000 features. Hence the PC1 and PC2 (%) that can describe the data is small.
#As we can see there is not a reliable way to know the total number of patient segments in exploratory analysis
PLOTPCA_PRCOMP(inputM=dtm,clus = as.character(KM_clus_cols$cluster) )

```

## Lets get some help from statistics to know the number of clusters
**Consensus cluster package allows us to discover the robust number of segments in an unsupervised setting **
```{r consensus clustering,echo=TRUE}

results = ConsensusClusterPlus(dtm,maxK=8,reps=100,pItem=0.8,pFeature=1,
                               clusterAlg="hc",distance="pearson",seed=25)

```
**This gives us a bunch of plots that shows how the cumulative distributive function and the area under the curve increases upto k=4 clusters beyond which it falls off; We can conclude that k=4 gives us robust number of clusters to work with going forward**
## Use Kaplan Meier plot to define clinically relevant population segments
``` {r KMplot, echo=TRUE}

clinical_patient_Cancer <- GDCquery_clinic("TCGA-BRCA","clinical")
 #clinical_patient_Cancer <- data.frame(
 #     bcr_patient_barcode = substr(colnms,1,12)
 #)
 
clinical_patient_Cancer1=clinical_patient_Cancer
clinical_patient_Cancer1$clus=as.vector(KM_clus_cols$cluster)[
                                (match(clinical_patient_Cancer$submitter_id,substr(colnms,1,12)))
                                ]
 
 
TCGAanalyze_survival(clinical_patient_Cancer1,
                     clusterCol="clus",
                     risk.table = F,
                     xlim = c(100,6000),
                     conf.int = FALSE,
                     color = c("Dark2"),
                     filename = NULL
                     )
 
```

## Since Cluster 4 does poor than Cluster 3. We can do statistics on genes (features) that is significantly associated with this cluster
**To do this, we use Kaplan-meier (KM) survival analysis - a univariate analysis; Each gene at a time is systematically used to separate patients with high/low expression and compared against the two groups of clusters**

```{r Genes for KMplot, echo=TRUE}

 
tabSurvKM <- TCGAanalyze_SurvivalKM(clinical_patient_Cancer,
                                     dtm,
                                     Genelist = rownames(dtm),
                                     Survresult = FALSE,
                                     p.cut = 0.4,
                                     ThreshTop = 0.7,
                                     ThreshDown = 0.2,
                                     group1 = names(KM_clus_cols$cluster==3), 
                                     group2 = names(KM_clus_cols$cluster==4)
                                    ) 

#Top 20 Genes - ranked by pvalue
tabSurvKM[1:20,]

 
```
